A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-24 tex/BiSig.tex
[ fix ] Substitution in (op) and the rule scheme

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically has a binary implication type $\bto$ and some base types so that the set of all types is at least inhabited.
  In our setting, the set of $\Sigma$-types over $\Theta$ may have variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is inhabited if $\Theta$ is.
  Hence, there is no need to introduce base types which are not introduced by any term construct.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus is thus given solely by an operation $\bto$ with $\arity(\bto) = 2$, i.e.\ a binary operation.
  For simply typed $\lambda$-calculus with binary products, we can augment
  $\Sigma_{\bto}$ by adding a binary operation $\btimes$ for the binary product type.
  We denote this augmented signature by $\Sigma_{\bto, \btimes}$.

\subsection{Binding Signatures and Intrinsically Typed Terms}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

\begin{example} \label{ex:STLC-sig}
  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
  \end{align*}
    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
  \begin{align*}
    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
    && 
    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
  \end{align*}
\end{example}

\begin{example}\label{ex:STLCx-sig}
  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
to $\Omega_{\bto}$:
  \begin{align*}
    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
    && \fst \colon A, B \vdash A \btimes B \to A
    && \snd \colon A, B \vdash A \btimes B \to B
  \end{align*}
\end{example}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma_{\Sigma, \Omega} \vdash A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho : \Xi \to \Theta \\\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
\end{figure}

\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\rho\colon \Xi \to \Theta \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
  \end{mathpar}
  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

>>>>

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
  In our setting, the set of $\Sigma$-types over $\Theta$ may contain variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is non-empty if $\Theta$ is.
  Therefore, there is no need to introduce base types that are not used by any term construct.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists solely of a binary operation $\bto$, where $\arity(\bto) = 2$.
  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
\end{example}

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

\begin{example} \label{ex:STLC-sig}
  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
  \end{align*}
    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
  \begin{align*}
    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
    && 
    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
  \end{align*}
  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
to $\Omega_{\bto}$:
  \begin{align*}
    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
    && \fst \colon A, B \vdash A \btimes B \to A
    && \snd \colon A, B \vdash A \btimes B \to B
  \end{align*}
\end{example}

First, it is noteworthy that the set of term constructs in a type theory need not be finite.
For instance, a type theory that includes ground types like arbitrary-bit integers can be specified by a nullary operation, i.e.\ a constant, for each integer.

More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
Instead of treating the application construct $\app_{A, B}$ as indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
This is not only for brevity but also necessary to incorporate the global context of type variables available, which is crucial for unification to solve constraints during type synthesis.

In contrast to the set $\Theta$ representing the context of global type variables for a derivation of a typing judgement $\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A$ introduced below, the set $\Xi$ associated with an operation can be referred to as the \emph{local context} for that operation.

\subsection{Intrinsically Typed Terms}

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp_o)$ for every operation $o$ in $\Omega$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Xi \to \Theta$ is used to instantiate explicitly variables in the local context $\Xi$ with global type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure that all types are will-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
  \end{mathpar}
  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}
